---
title: "MTH6139 Time Series"
subtitle: "Coursework 1 – Prophet Forecasting Project"
author: "Muhammad Taha Khan"
date: "March 12, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
    highlight: tango
---

# 1 Introduction

The aim of this project is to analyze a given time series data set and forecast it using the Prophet Forecasting System developed by Meta.

The data set used in this project is the built-in data set **co2** available in R. The data set represents monthly atmospheric carbon dioxide concentrations observed at Mauna Loa Observatory in Hawaii.

---

# 2 Exploring the Dataset

```{r}
data(co2)

start(co2)
end(co2)
frequency(co2)
```

The dataset begins in 1959 and contains monthly observations of atmospheric CO₂ levels.

---

# 3 Visualising the Time Series

```{r}
plot(co2,
     main = "Atmospheric CO2 Concentration",
     xlab = "Year",
     ylab = "CO2 (ppm)")
```

The plot shows a clear upward trend in atmospheric carbon dioxide levels along with a repeating seasonal pattern.

---

# 4 Preparing the Data for Prophet

```{r}
library(prophet)
library(zoo)

time_values <- time(co2)
date_values <- as.Date(as.yearmon(time_values))

co2_dataframe <- data.frame(
  ds = date_values,
  y = as.numeric(co2)
)

head(co2_dataframe)
```

Prophet requires the data to be stored in a dataframe with two columns: **ds** (date) and **y** (value).

---

# 5 Building the Prophet Model

```{r}
prophet_model <- prophet(co2_dataframe)
```

Prophet automatically detects long-term trends and seasonal patterns.

---

# 6 Forecasting Future Values

```{r}
future_dates <- make_future_dataframe(prophet_model,
                                      periods = 24,
                                      freq = "month")

forecast_results <- predict(prophet_model, future_dates)

head(forecast_results)
```

The model forecasts CO₂ levels for the next 24 months.

---

# 7 Forecast Plot

```{r}
plot(prophet_model, forecast_results)
```

The dark line represents predicted values, while the shaded region represents uncertainty intervals.

---

# 8 Model Components

```{r}
prophet_plot_components(prophet_model, forecast_results)
```

This shows the trend and seasonal components of the model.

---

# 9 Conclusion

This project used the Prophet forecasting system to analyze the atmospheric CO₂ data. The data has a strong upward trend and clear yearly seasonality. The forecast indicates that CO₂ levels will continue to rise if the current trends persist.

---

# References

Prophet Documentation  
https://facebook.github.io/prophet/

R Markdown Documentation  
https://rmarkdown.rstudio.com